
TERRE Differential DNAm Analysis
===============================================================

##### Analyst: Samantha Schaffner
##### Date: Oct 20, 2023
  
The *TERRE* cohort (subset containing n=88 cases and n=177 controls) consists of French agricultural workers enrolled in the Mutualite Sociale Agricole health insurance system. All PD cases in this subset of data have a disease duration of one year or less. 

In this script I will perform EWAS in TERRE using the blood reference CMRs from the EPIC array, with median betas calculated within each sex of TERRE (see ~/2-EWAS_CMRs/TERRE/TERRE_models.Rmd). I will take random subsets of males and females, to perform a null EWAS. This will reveal how many CMRs are picked up by chance in differential methylation analysis using these sample sizes.

## Loading Libraries
```{r libraries, eval=F}
setwd("/home1/NEURO/schaffner/")
library(dplyr)
library(MASS)
library(sfsmisc)
source("~/2-EWAS_CMRs/TERRE/sensitivity_analyses/CMR_rlm_perm.R")
```


# Null EWAS with sexes stratified

## Females

```{r rlm F, eval = F}
meta_F <- read.csv("~/1-preprocessing/TERRE/propensity_matching/terre_meta_F_probit_matched_4.5.3.csv")
load("~/2-EWAS_CMRs/TERRE/refCMR_medB_F_var.RData")
all(meta_F$patient==rownames(refCMR_medB_F_var)) #TRUE

#check formatting of meta data
str(meta_F[,c("age","smoking","head_trauma","alcohol1","CTP_PC1","CTP_PC2","CTP_PC3","CTP_PC4","CTP_PC5","CTP_PC6","ancestryPC1","ancestryPC2","ancestryPC3","plate","SentrixPosition_A")])
meta_F$smoking <- as.factor(meta_F$smoking)
meta_F$head_trauma <- as.factor(meta_F$head_trauma)
meta_F$alcohol1 <- as.factor(meta_F$alcohol1)
meta_F$plate <- as.factor(meta_F$plate)

#run permutation
#this will shuffle the PD labels on every iteration, running the full EWAS model without weights
#p-value is returned for whether the number of DM-CMRs (adjDB>0.03, padj<0.05) is greater or less than expected by change
#system.time(perm <- CMR_rlm_perm(refCMR_medB_F_var, meta_F, nDM_CMRs=69, nperm=10)) #1 hour per 10 permutations
#[1] "Permutation P values"
#[1] "Enrichment: 0; Depletion 1"

#perm <- CMR_rlm_perm(refCMR_medB_F_var, meta_F, nDM_CMRs=69, nperm=100)
#[1] "Permutation P values"
#[1] "Enrichment: 0; Depletion 1"

perm_f <- CMR_rlm_perm(refCMR_medB_F_var, meta_F, nDM_CMRs=69, nperm=1000)
#[1] "Permutation P values"
#[1] "Enrichment: 0.006; Depletion 0.994"
```

## Males

```{r rlm F, eval = F}
meta_M <- read.csv("~/1-preprocessing/TERRE/propensity_matching/terre_meta_M_probit_matched_4.5.3.csv")
load("~/2-EWAS_CMRs/TERRE/refCMR_medB_M_var.RData")
all(meta_M$patient==rownames(refCMR_medB_M_var)) #TRUE

#check formatting of meta data
str(meta_M[,c("age","smoking","head_trauma","alcohol1","CTP_PC1","CTP_PC2","CTP_PC3","CTP_PC4","CTP_PC5","CTP_PC6","ancestryPC1","ancestryPC2","ancestryPC3","plate","SentrixPosition_A")])
meta_M$smoking <- as.factor(meta_M$smoking)
meta_M$head_trauma <- as.factor(meta_M$head_trauma)
meta_M$alcohol1 <- as.factor(meta_M$alcohol1)
meta_M$plate <- as.factor(meta_M$plate)

#run permutation
#this will shuffle the PD labels on every iteration, running the full EWAS model without weights
#p-value is returned for whether the number of DM-CMRs (adjDB>0.03, padj<0.05) is greater or less than expected by change
#perm <- CMR_rlm_perm(refCMR_medB_M_var, meta_M, nDM_CMRs=2, nperm=100)
#[1] "Permutation P values"
#[1] "Enrichment: 0.06; Depletion 1"

perm_m <- CMR_rlm_perm(refCMR_medB_M_var, meta_M, nDM_CMRs=2, nperm=1000)
#"Enrichment: 0.022; Depletion 0.984"
```